---
title: "Parent W3 Scored"
author: "Shakil Rafi"
format: html
toc: true
toc-depth: 4
lightbox: true
title-block-banner: true
df-print: tibble
comments:
  hypothesis: true
editor: visual
---

## Read Data

```{r}
library(tidyverse)
data <- read_csv("parent_wave_3_scored.csv")
```

## EDA

### Geographic Distribution and DOB

```{r}
#| eval: false
library(tidyverse)
library(tidygeocoder)
library(maps)

# Step 1: Clean the p3cpob field
data_clean <- data |>
  filter(!is.na(p3cpob)) |>
  mutate(p3cpob_clean = str_to_title(p3cpob)) |>
  mutate(p3cpob_clean = str_replace_all(p3cpob_clean, "\\s+", " "),
         p3cpob_clean = str_remove_all(p3cpob_clean, "\\s+Usa|\\s+Us|\\s+Usa\\.?|\\s+Us\\.?"),
         p3cpob_clean = str_remove_all(p3cpob_clean, ",\\s*$")) |>
  mutate(p3cpob_clean = str_trim(p3cpob_clean)) 

# Step 2: Extract unique cleaned locations
unique_locations <- data_clean |>
  distinct(p3cpob_clean)

# Step 3: Geocode
geocoded <- unique_locations |>
  geocode(p3cpob_clean, method = "osm", lat = latitude , long = longitude)

# Step 4: Merge geocodes back into the main data
data_geo <- data_clean |>
  left_join(geocoded, by = "p3cpob_clean")

# Step 5: Plot using ggplot2
us_map <- map_data("world")

us_map |>
ggplot() +
  geom_polygon(aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "white") +
  geom_point(data = data_geo, aes(x = longitude, y = latitude), 
             color = "red", alpha = 0.7, size = 2) +
  coord_fixed(1.3) +
  theme_minimal() +
  labs(title = "Place of Birth (p3cpob)", x = "Longitude", y = "Latitude")

```

```{r}
library(tidyverse)
library(lubridate)

# Step 1: Parse p3cdob as a date
data <- data %>%
  mutate(p3cdob = parse_date(p3cdob, format = "%m/%d/%Y"))  # Adjust format if necessary

# Step 2: Extract year of birth
data <- data %>%
  filter(!is.na(p3cdob)) %>%
  mutate(birth_year = year(p3cdob))

# Step 3: Plot as a barplot
data |>
ggplot(aes(x = birth_year)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(
    title = "Distribution of Child Date of Birth (p3cdob)",
    x = "Year of Birth",
    y = "Count"
  )

```

### Descriptive Statistics

```{r}
data |> 
  select(where(is.numeric)) |>
  psych::describe()
```

### Normality Testing

```{r}
library(ggplot2)
library(reshape2)
library(dplyr)
library(RColorBrewer)

# Load data
data <- read.csv("parent_wave_3_scored.csv")

# Psychometric variables
psychometrics <- c(
  "p3sp", "p3axdp", "p3wtd", "p3rbb", "p3agg", "p3smc",
  "p3thpr", "p3atn", "p3pix", "p3ppx", "p3pmx"
)

# Subset and remove rows with NAs
df <- data %>%
  select(all_of(psychometrics)) %>%
  na.omit()

# Compute correlation matrix
cor_matrix <- cor(df, use = "pairwise.complete.obs")

# Melt into long format
cor_melted <- melt(cor_matrix)

# Plot heatmap
ggplot(cor_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "blue", high = "red", mid = "white",
    midpoint = 0, limit = c(-1, 1), space = "Lab",
    name = "Correlation"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(
    title = "Correlation Heatmap of Psychometric Scores",
    x = "", y = ""
  )

```

```{r}
library(ggplot2)
library(dplyr)
library(scales)  # for hue_pal

# Load your data
psychometrics <- c(
  "p3sp", "p3axdp", "p3wtd", "p3rbb", "p3agg", "p3smc",
  "p3thpr", "p3atn", "p3pix", "p3ppx", "p3pmx"
)

# Psychometric variables with human-readable names
psychometrics <- c(
  p3sp   = "Sleep Problems",
  p3axdp = "Anxiety/Depression",
  p3wtd  = "Withdrawn/Depressed",
  p3rbb  = "Rule-Breaking Behavior",
  p3agg  = "Aggressive Behavior",
  p3smc  = "Somatic Complaints",
  p3thpr = "Thought Problems",
  p3atn  = "Attention Problems",
  p3pix  = "Parent Internalizing Index",
  p3ppx  = "Parent Externalizing Index",
  p3pmx  = "Parent Total Problems Index"
)

# Generate distinct pastel colors
colors <- scales::hue_pal(h = c(0, 360) + 15, c = 30, l = 90)(length(psychometrics))

# Loop through each variable
i <- 1
for (var in names(psychometrics)) {
  label <- psychometrics[[var]]
  p <- ggplot(data, aes(x = .data[[var]])) +
    geom_density(fill = colors[i], alpha = 0.7, na.rm = TRUE) +
    labs(
      title = paste("Density of", label, "(", var,")"),
      x = label,
      y = "Density"
    ) +
    theme_minimal(base_size = 14)
  print(p)
  i <- i + 1
}

```

#### Shapiro-Wilk

```{r}
results <- lapply(names(psychometrics), function(var) {
  values <- data[[var]]
  values <- values[!is.na(values)]
  if (length(values) >= 3 && length(values) <= 5000) {
    test <- shapiro.test(values)
    tibble(
      Variable = psychometrics[[var]],
      W_statistic = test$statistic,
      p_value = test$p.value
    )
  } else {
    tibble(
      Variable = psychometrics[[var]],
      W_statistic = NA,
      p_value = NA
    )
  }
})

# Combine into a data frame
shapiro_results <- bind_rows(results)

# Display results
print(shapiro_results)
```

### Parent Education Levels Analysis

```{r}
library(ggplot2)

# Load data
df <- read.csv("parent_wave_3_scored.csv")

# Variables to plot with human-friendly titles
plot_vars <- list(
  p3sp   = "Sleep Problems",
  p3axdp = "Anxiety/Depression",
  p3wtd  = "Withdrawn/Depressed",
  p3rbb  = "Rule-Breaking Behavior",
  p3agg  = "Aggressive Behavior",
  p3smc  = "Somatic Complaints",
  p3thpr = "Thought Problems",
  p3atn  = "Attention Problems",
  p3pix  = "Parent Internalizing Index",
  p3ppx  = "Parent Externalizing Index",
  p3pmx  = "Parent Total Problems Index"
)

# Loop through and generate pastel violin plots without legend
for (var in names(plot_vars)) {
  title <- plot_vars[[var]]
  p <- ggplot(df, aes(x = as.factor(p3edu1), y = .data[[var]], fill = as.factor(p3edu1))) +
    geom_violin(trim = FALSE) +
    geom_jitter(width = 0.2, alpha = 0.3, color = "black") +
    scale_fill_brewer(palette = "Pastel1") +
    labs(
      title = paste(title, "by Parent 1 Edu Level (", var, ")"),
      x = "Parent 1 Education Level",
      y = title
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")
  print(p)
}

```

```{r}
library(ggplot2)

# Load data
df <- read.csv("parent_wave_3_scored.csv")

# Variables to plot with human-friendly titles
plot_vars <- list(
  p3sp   = "Sleep Problems",
  p3axdp = "Anxiety/Depression",
  p3wtd  = "Withdrawn/Depressed",
  p3rbb  = "Rule-Breaking Behavior",
  p3agg  = "Aggressive Behavior",
  p3smc  = "Somatic Complaints",
  p3thpr = "Thought Problems",
  p3atn  = "Attention Problems",
  p3pix  = "Parent Internalizing Index",
  p3ppx  = "Parent Externalizing Index",
  p3pmx  = "Parent Total Problems Index"
)

# Loop through and generate pastel violin plots without legend
for (var in names(plot_vars)) {
  title <- plot_vars[[var]]
  p <- ggplot(df, aes(x = as.factor(p3edu2), y = .data[[var]], fill = as.factor(p3edu2))) +
    geom_violin(trim = FALSE) +
    geom_jitter(width = 0.2, alpha = 0.3, color = "black") +
    scale_fill_brewer(palette = "Pastel2") +
    labs(
      title = paste(title, "by Parent 2 Edu Level (",var,")"),
      x = "Parent 2 Education Level",
      y = title
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")
  print(p)
}
```

### Parent Marital Status

```{r}
library(ggplot2)

# Load data
df <- read.csv("parent_wave_3_scored.csv")

# Variables to plot with human-friendly titles
plot_vars <- list(
  p3sp   = "Sleep Problems",
  p3axdp = "Anxiety/Depression",
  p3wtd  = "Withdrawn/Depressed",
  p3rbb  = "Rule-Breaking Behavior",
  p3agg  = "Aggressive Behavior",
  p3smc  = "Somatic Complaints",
  p3thpr = "Thought Problems",
  p3atn  = "Attention Problems",
  p3pix  = "Parent Internalizing Index",
  p3ppx  = "Parent Externalizing Index",
  p3pmx  = "Parent Total Problems Index"
)

# Loop through and generate pastel violin plots without legend
for (var in names(plot_vars)) {
  title <- plot_vars[[var]]
  p <- ggplot(df, aes(x = as.factor(p3mar1), y = .data[[var]], fill = as.factor(p3mar1))) +
    geom_violin(trim = FALSE) +
    geom_jitter(width = 0.2, alpha = 0.3, color = "black") +
    scale_fill_brewer(palette = "Pastel1") +
    labs(
      title = paste(title, "by Mar Status (", var, ")"),
      x = "Marital Status",
      y = title
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")
  print(p)
}
```

### ANOVA and Effect Size

#### ANOVA vs p3edu

```{r}

# Psychometric variables (as before)
psychometrics <- c(
  p3sp   = "Sleep Problems",
  p3axdp = "Anxiety/Depression",
  p3wtd  = "Withdrawn/Depressed",
  p3rbb  = "Rule-Breaking Behavior",
  p3agg  = "Aggressive Behavior",
  p3smc  = "Somatic Complaints",
  p3thpr = "Thought Problems",
  p3atn  = "Attention Problems",
  p3pix  = "Parent Internalizing Index",
  p3ppx  = "Parent Externalizing Index",
  p3pmx  = "Parent Total Problems Index"
)

# Loop through and print ANOVA summary for each
for (var in names(psychometrics)) {
  cat("\n--------------------------------------------\n")
  cat("ANOVA for", psychometrics[[var]], "(", var, ")\n")
  cat("--------------------------------------------\n")
  
  result <- summary(aov(as.formula(paste(var, "~ as.factor(p3edu1)")), data = data))
  print(result)
}

```

```{r}

# Psychometric variables (as before)
psychometrics <- c(
  p3sp   = "Sleep Problems",
  p3axdp = "Anxiety/Depression",
  p3wtd  = "Withdrawn/Depressed",
  p3rbb  = "Rule-Breaking Behavior",
  p3agg  = "Aggressive Behavior",
  p3smc  = "Somatic Complaints",
  p3thpr = "Thought Problems",
  p3atn  = "Attention Problems",
  p3pix  = "Parent Internalizing Index",
  p3ppx  = "Parent Externalizing Index",
  p3pmx  = "Parent Total Problems Index"
)

# Loop through and print ANOVA summary for each
for (var in names(psychometrics)) {
  cat("\n--------------------------------------------\n")
  cat("ANOVA for", psychometrics[[var]], "(", var, ")\n")
  cat("--------------------------------------------\n")
  
  result <- summary(aov(as.formula(paste(var, "~ as.factor(p3edu2)")), data = data))
  print(result)
}
```

#### ANOVA vs p3mar1

```{r}
# Psychometric variables (as before)
psychometrics <- c(
  p3sp   = "Sleep Problems",
  p3axdp = "Anxiety/Depression",
  p3wtd  = "Withdrawn/Depressed",
  p3rbb  = "Rule-Breaking Behavior",
  p3agg  = "Aggressive Behavior",
  p3smc  = "Somatic Complaints",
  p3thpr = "Thought Problems",
  p3atn  = "Attention Problems",
  p3pix  = "Parent Internalizing Index",
  p3ppx  = "Parent Externalizing Index",
  p3pmx  = "Parent Total Problems Index"
)

# Loop through and print ANOVA summary for each
for (var in names(psychometrics)) {
  cat("\n--------------------------------------------\n")
  cat("ANOVA for", psychometrics[[var]], "(", var, ")\n")
  cat("--------------------------------------------\n")
  
  result <- summary(aov(as.formula(paste(var, "~ as.factor(p3mar1)")), data = data))
  print(result)
}
```

#### ANOVA vs p3emp

```{r}
# Psychometric variables (as before)
psychometrics <- c(
  p3sp   = "Sleep Problems",
  p3axdp = "Anxiety/Depression",
  p3wtd  = "Withdrawn/Depressed",
  p3rbb  = "Rule-Breaking Behavior",
  p3agg  = "Aggressive Behavior",
  p3smc  = "Somatic Complaints",
  p3thpr = "Thought Problems",
  p3atn  = "Attention Problems",
  p3pix  = "Parent Internalizing Index",
  p3ppx  = "Parent Externalizing Index",
  p3pmx  = "Parent Total Problems Index"
)

# Loop through and print ANOVA summary for each
for (var in names(psychometrics)) {
  cat("\n--------------------------------------------\n")
  cat("ANOVA for", psychometrics[[var]], "(", var, ")\n")
  cat("--------------------------------------------\n")
  
  result <- summary(aov(as.formula(paste(var, "~ as.factor(p3emp1)")), data = data))
  print(result)
}
```

```{r}
# Psychometric variables (as before)
psychometrics <- c(
  p3sp   = "Sleep Problems",
  p3axdp = "Anxiety/Depression",
  p3wtd  = "Withdrawn/Depressed",
  p3rbb  = "Rule-Breaking Behavior",
  p3agg  = "Aggressive Behavior",
  p3smc  = "Somatic Complaints",
  p3thpr = "Thought Problems",
  p3atn  = "Attention Problems",
  p3pix  = "Parent Internalizing Index",
  p3ppx  = "Parent Externalizing Index",
  p3pmx  = "Parent Total Problems Index"
)

# Loop through and print ANOVA summary for each
for (var in names(psychometrics)) {
  cat("\n--------------------------------------------\n")
  cat("ANOVA for", psychometrics[[var]], "(", var, ")\n")
  cat("--------------------------------------------\n")
  
  result <- summary(aov(as.formula(paste(var, "~ as.factor(p3emp2)")), data = data))
  print(result)
}
```

#### Effect Size and Kruskal-Wallis vs p3edu

```{r}
library(effectsize)

# Load data
data <- read.csv("parent_wave_3_scored.csv")

# Psychometric variables
psychometrics <- c(
  "p3sp", "p3axdp", "p3wtd", "p3rbb", "p3agg", "p3smc",
  "p3thpr", "p3atn", "p3pix", "p3ppx", "p3pmx"
)

# Loop: run ANOVA and compute eta squared
for (var in psychometrics) {
  formula <- as.formula(paste(var, "~ as.factor(p3edu1)"))
  
  # Run ANOVA
  model <- aov(formula, data = data)
  
  # Eta Squared
  eta <- eta_squared(model, partial = FALSE)
  
  # Kruskal-Wallis
  kw_p <- kruskal.test(formula, data = data)$p.value
  
  cat("\n----------------------------------------\n")
  cat("Results for", var, "\n")
  cat("----------------------------------------\n")
  print(eta)
  cat("Kruskal-Wallis p-value:", round(kw_p, 4), "\n")
}

```

```{r}
library(effectsize)

# Load data
data <- read.csv("parent_wave_3_scored.csv")

# Psychometric variables
psychometrics <- c(
  "p3sp", "p3axdp", "p3wtd", "p3rbb", "p3agg", "p3smc",
  "p3thpr", "p3atn", "p3pix", "p3ppx", "p3pmx"
)

# Loop: run ANOVA and compute eta squared
for (var in psychometrics) {
  formula <- as.formula(paste(var, "~ as.factor(p3edu2)"))
  
  # Run ANOVA
  model <- aov(formula, data = data)
  
  # Eta Squared
  eta <- eta_squared(model, partial = FALSE)
  
  # Kruskal-Wallis
  kw_p <- kruskal.test(formula, data = data)$p.value
  
  cat("\n----------------------------------------\n")
  cat("Results for", var, "\n")
  cat("----------------------------------------\n")
  print(eta)
  cat("Kruskal-Wallis p-value:", round(kw_p, 4), "\n")
}
```

#### Effect Size and Kruskal-Wallis vs p3mar1

```{r}
library(effectsize)

# Load data
data <- read.csv("parent_wave_3_scored.csv")

# Psychometric variables
psychometrics <- c(
  "p3sp", "p3axdp", "p3wtd", "p3rbb", "p3agg", "p3smc",
  "p3thpr", "p3atn", "p3pix", "p3ppx", "p3pmx"
)

# Loop: run ANOVA and compute eta squared
for (var in psychometrics) {
  formula <- as.formula(paste(var, "~ as.factor(p3mar1)"))
  
  # Run ANOVA
  model <- aov(formula, data = data)
  
  # Eta Squared
  eta <- eta_squared(model, partial = FALSE)
  
  # Kruskal-Wallis
  kw_p <- kruskal.test(formula, data = data)$p.value
  
  cat("\n----------------------------------------\n")
  cat("Results for", var, "\n")
  cat("----------------------------------------\n")
  print(eta)
  cat("Kruskal-Wallis p-value:", round(kw_p, 4), "\n")
}
```

#### Effect size and Kruskal-Wallis vs p3emp

```{r}
library(effectsize)

# Load data
data <- read.csv("parent_wave_3_scored.csv")

# Psychometric variables
psychometrics <- c(
  "p3sp", "p3axdp", "p3wtd", "p3rbb", "p3agg", "p3smc",
  "p3thpr", "p3atn", "p3pix", "p3ppx", "p3pmx"
)

# Loop: run ANOVA and compute eta squared
for (var in psychometrics) {
  formula <- as.formula(paste(var, "~ as.factor(p3emp1)"))
  
  # Run ANOVA
  model <- aov(formula, data = data)
  
  # Eta Squared
  eta <- eta_squared(model, partial = FALSE)
  
  # Kruskal-Wallis
  kw_p <- kruskal.test(formula, data = data)$p.value
  
  cat("\n----------------------------------------\n")
  cat("Results for", var, "\n")
  cat("----------------------------------------\n")
  print(eta)
  cat("Kruskal-Wallis p-value:", round(kw_p, 4), "\n")
}
```

```{r}
library(effectsize)

# Load data
data <- read.csv("parent_wave_3_scored.csv")

# Psychometric variables
psychometrics <- c(
  "p3sp", "p3axdp", "p3wtd", "p3rbb", "p3agg", "p3smc",
  "p3thpr", "p3atn", "p3pix", "p3ppx", "p3pmx"
)

# Loop: run ANOVA and compute eta squared
for (var in psychometrics) {
  formula <- as.formula(paste(var, "~ as.factor(p3emp2)"))
  
  # Run ANOVA
  model <- aov(formula, data = data)
  
  # Eta Squared
  eta <- eta_squared(model, partial = FALSE)
  
  # Kruskal-Wallis
  kw_p <- kruskal.test(formula, data = data)$p.value
  
  cat("\n----------------------------------------\n")
  cat("Results for", var, "\n")
  cat("----------------------------------------\n")
  print(eta)
  cat("Kruskal-Wallis p-value:", round(kw_p, 4), "\n")
}
```

## General Linear Model

```{r}
# Load data
data <- read.csv("parent_wave_3_scored.csv")

# Psychometric outcome variables
outcomes <- c(
  "p3sp", "p3axdp", "p3wtd", "p3rbb", "p3agg", "p3smc",
  "p3thpr", "p3atn", "p3pix", "p3ppx", "p3pmx"
)

# Predictor variables (example)
predictors <- c("p3edu1", "p3incom", "p3pgen", "p3page", "p3mar1", "p3emp1", "p3chol1")

# Loop through and fit model
for (outcome in outcomes) {
  formula <- as.formula(paste(outcome, "~", paste(predictors, collapse = " + ")))
  model <- lm(formula, data = data)
  
  cat("\n===========================================\n")
  cat("Linear Model Summary for", outcome, "\n")
  cat("===========================================\n")
  print(summary(model))
}

```

```{r}
library(randomForest)
library(dplyr)
library(ggplot2)

# Load data
data <- read.csv("parent_wave_3_scored.csv")

# Psychometric outcomes with human-readable names
psychometrics <- c(
  p3sp   = "Sleep Problems",
  p3axdp = "Anxiety/Depression",
  p3wtd  = "Withdrawn/Depressed",
  p3rbb  = "Rule-Breaking Behavior",
  p3agg  = "Aggressive Behavior",
  p3smc  = "Somatic Complaints",
  p3thpr = "Thought Problems",
  p3atn  = "Attention Problems",
  p3pix  = "Parent Internalizing Index",
  p3ppx  = "Parent Externalizing Index",
  p3pmx  = "Parent Total Problems Index"
)

# Predictor variable labels (customize these based on your data dictionary)
predictor_labels <- c(
  p3edu1  = "Education Level",
  p3incom = "Household Income",
  p3pgen  = "Parent Gender",
  p3page  = "Parent Age",
  p3mar1  = "Marital Status of Parent 1",
  p3emp1  = "Employment Status of Parent 1",
  p3chol1 = "Child Health Insurance"
)

# Convert character predictors to factors
data <- data %>%
  mutate(across(all_of(names(predictor_labels)), ~ if (is.character(.x) || is.factor(.x)) as.factor(.x) else .x))

# Loop through each psychometric variable
for (var in names(psychometrics)) {
  outcome_label <- psychometrics[[var]]
  formula <- as.formula(paste(var, "~", paste(names(predictor_labels), collapse = " + ")))
  df <- data[, c(var, names(predictor_labels))] %>% na.omit()
  
  rf <- randomForest(formula, data = df, importance = TRUE)
  importance_df <- as.data.frame(importance(rf))
  importance_df$Variable <- rownames(importance_df)
  
  # Replace predictor names with human-readable labels
  importance_df <- importance_df %>%
    mutate(Predictor_Label = paste0(predictor_labels[Variable], " (", Variable, ")")) %>%
    arrange(desc(`%IncMSE`))
  
  # Plot
  ggplot(importance_df, aes(x = reorder(Predictor_Label, `%IncMSE`), y = `%IncMSE`)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    labs(
      title = paste("Variable Importance for", outcome_label, "(", var, ")"),
      x = "Predictor",
      y = "% Increase in MSE (Importance)"
    ) +
    theme_minimal(base_size = 14) -> p
  
  print(p)
}

```
