---
title: "W1 Parent Scored"
author: "Shakil Rafi"
format: html
toc: true
toc-depth: 4
lightbox: true
title-block-banner: true
df-print: tibble
comments:
  hypothesis: true
editor: visual
---

## Read Data

```{r}
library(tidyverse)

data <- read_csv("w1-parent-scored.csv")
```

```{r}
#| eval: false
library(tidygeocoder)
library(dplyr)
library(ggplot2)
library(maps)
library(readr)

# Load your CSV
df <- read_csv("w1-parent-scored.csv")

# Drop missing or blank entries
locations <- df %>%
  select(p1cpob) %>%
  filter(!is.na(p1cpob), p1cpob != "") %>%
  distinct()

# Geocode using OpenStreetMap Nominatim (free and no API key required)
geocoded <- locations %>%
  geocode(p1cpob, method = "osm", lat = latitude, long = longitude)

# Drop failed lookups
geocoded_clean <- geocoded %>% filter(!is.na(latitude) & !is.na(longitude))

# Plot on a world map
world_map <- map_data("world")

ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group),
               fill = "lightgray", color = "white") +
  geom_point(data = geocoded_clean, aes(x = longitude, y = latitude),
             color = "red", size = 2, alpha = 0.7) +
  coord_fixed(1.3) +
  labs(title = "Locations from p1cpob", x = "Longitude", y = "Latitude") +
  theme_minimal()

```

```{r}
data |>
  psych::describe()
```

## EDA

### Descriptive Statistics

```{r}
library(ggplot2)
library(dplyr)
library(tibble)
library(readr)

# Load the data
df <- read_csv("w1-parent-scored.csv")
df <- df %>% select(-SID)

# Pastel palette
pastel_colors <- c("#AEC6CF", "#FFB347", "#B39EB5", "#77DD77", "#F49AC2")

# Loop through columns and plot
for (i in seq_along(df)) {
  col_name <- names(df)[i]
  color <- pastel_colors[(i - 1) %% length(pastel_colors) + 1]
  
  p <- ggplot(df, aes_string(x = col_name)) +
    geom_density(fill = color, alpha = 0.7, color = "black", linewidth = 0.5) +  # Eyeliner, baby
    labs(title = paste("Density Plot of", col_name)) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5),
      axis.title.y = element_blank()
    )
  
  print(p)
}
```

```{r}
shapiro_results <- df |>
    select(where(is.numeric))|>
  summarise(across(everything(), ~ list(shapiro.test(.x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "test") |>
  mutate(
    statistic = map_dbl(test, ~ .x$statistic),
    p_value = map_dbl(test, ~ .x$p.value)
  ) |>
  select(variable, statistic, p_value)

shapiro_results
```

```{r, fig.width=12, fig.height=12}

library(tidyverse)

# Load your file
df <- read_csv("w1-parent-scored.csv")

# Compute correlations on numeric columns
corr <- df |>
  select(where(is.numeric)) |>
  cor(use = "pairwise.complete.obs")  # safer for missing data

# Convert to tidy format for plotting
corr_df <- as.data.frame(as.table(corr))

# Plot heatmap
ggplot(corr_df, aes(Var1, Var2, fill = Freq)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(
    low = "#d0d1e6",
    mid = "#f7f7f7",
    high = "#1c9099",
    midpoint = 0,
    name = "Correlation"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 12),
    axis.text.y = element_text(size = 12),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16)
  ) +
  labs(
    title = "Correlation Heatmap",
    x = NULL,
    y = NULL
  ) +
  coord_fixed()


```

## Parent Education Level Analysis

```{r}
library(ggplot2)

# Load data
df <- read.csv("w1-parent-scored.csv")

# Variables to plot with human-friendly titles
plot_vars <- list(
  p1sp   = "Parental Support",
  p1anxdp = "Anxiety/Depression",
  p1wtd  = "Weighted Stress",
  p1rbb  = "Risky Behavior Beliefs",
  p1agg  = "Aggression",
  p1smc  = "Social Media Concerns",
  p1thpr = "Thought Problems",
  p1atnpr = "Attention Problems",
  p1pix  = "Parental Involvement Index",
  p1ppx  = "Parental External Stress",
  p1pmsx = "Monitoring Social Experience"
)


# Loop through and generate pastel violin plots without legend
for (var in names(plot_vars)) {
  title <- plot_vars[[var]]
  p <- ggplot(df, aes(x = as.factor(p1edu1), y = .data[[var]], fill = as.factor(p1edu1))) +
    geom_violin(trim = FALSE) +
    geom_jitter(width = 0.2, alpha = 0.3, color = "black") +
    scale_fill_brewer(palette = "Pastel1") +
    labs(
      title = paste(title, "by Parent 1 Edu Level (", var, ")"),
      x = "Parent 1 Education Level",
      y = title
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")
  print(p)
}

```

```{r}
library(ggplot2)

# Load data
df <- read.csv("w1-parent-scored.csv")

# Variables to plot with human-friendly titles
plot_vars <- list(
  p1sp   = "Parental Support",
  p1anxdp = "Anxiety/Depression",
  p1wtd  = "Weighted Stress",
  p1rbb  = "Risky Behavior Beliefs",
  p1agg  = "Aggression",
  p1smc  = "Social Media Concerns",
  p1thpr = "Thought Problems",
  p1atnpr = "Attention Problems",
  p1pix  = "Parental Involvement Index",
  p1ppx  = "Parental External Stress",
  p1pmsx = "Monitoring Social Experience"
)


# Loop through and generate pastel violin plots without legend
for (var in names(plot_vars)) {
  title <- plot_vars[[var]]
  p <- ggplot(df, aes(x = as.factor(p1edu2), y = .data[[var]], fill = as.factor(p1edu2))) +
    geom_violin(trim = FALSE) +
    geom_jitter(width = 0.2, alpha = 0.3, color = "black") +
    scale_fill_brewer(palette = "Pastel1") +
    labs(
      title = paste(title, "by Parent 2 Edu Level (", var, ")"),
      x = "Parent 1 Education Level",
      y = title
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")
  print(p)
}

```

## Parent Marital Status Analysis

```{r}
library(ggplot2)

# Load data
df <- read.csv("w1-parent-scored.csv")

# Variables to plot with human-friendly titles
plot_vars <- list(
  p1sp   = "Parental Support",
  p1anxdp = "Anxiety/Depression",
  p1wtd  = "Weighted Stress",
  p1rbb  = "Risky Behavior Beliefs",
  p1agg  = "Aggression",
  p1smc  = "Social Media Concerns",
  p1thpr = "Thought Problems",
  p1atnpr = "Attention Problems",
  p1pix  = "Parental Involvement Index",
  p1ppx  = "Parental External Stress",
  p1pmsx = "Monitoring Social Experience"
)

# Loop through and generate pastel violin plots without legend
for (var in names(plot_vars)) {
  title <- plot_vars[[var]]
  p <- ggplot(df, aes(x = as.factor(p1mar1), y = .data[[var]], fill = as.factor(p1mar1))) +
    geom_violin(trim = FALSE) +
    geom_jitter(width = 0.2, alpha = 0.3, color = "black") +
    scale_fill_brewer(palette = "Pastel1") +
    labs(
      title = paste(title, "by Mar Status (", var, ")"),
      x = "Marital Status",
      y = title
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")
  print(p)
}
```

## Parent Employment Status Analysis

```{r}
library(ggplot2)

# Load data
df <- read.csv("w1-parent-scored.csv")

# Variables to plot with human-friendly titles
plot_vars <- list(
  p1sp   = "Parental Support",
  p1anxdp = "Anxiety/Depression",
  p1wtd  = "Weighted Stress",
  p1rbb  = "Risky Behavior Beliefs",
  p1agg  = "Aggression",
  p1smc  = "Social Media Concerns",
  p1thpr = "Thought Problems",
  p1atnpr = "Attention Problems",
  p1pix  = "Parental Involvement Index",
  p1ppx  = "Parental External Stress",
  p1pmsx = "Monitoring Social Experience"
)

# Loop through and generate pastel violin plots without legend
for (var in names(plot_vars)) {
  title <- plot_vars[[var]]
  p <- ggplot(df, aes(x = as.factor(p1emp1), y = .data[[var]], fill = as.factor(p1emp1))) +
    geom_violin(trim = FALSE) +
    geom_jitter(width = 0.2, alpha = 0.3, color = "black") +
    scale_fill_brewer(palette = "Pastel1") +
    labs(
      title = paste(title, "by Parent 1 Emp Status (", var, ")"),
      x = "Emp Status",
      y = title
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")
  print(p)
}
```

```{r}
library(ggplot2)

# Load data
df <- read.csv("w1-parent-scored.csv")

# Variables to plot with human-friendly titles
plot_vars <- list(
  p1sp   = "Parental Support",
  p1anxdp = "Anxiety/Depression",
  p1wtd  = "Weighted Stress",
  p1rbb  = "Risky Behavior Beliefs",
  p1agg  = "Aggression",
  p1smc  = "Social Media Concerns",
  p1thpr = "Thought Problems",
  p1atnpr = "Attention Problems",
  p1pix  = "Parental Involvement Index",
  p1ppx  = "Parental External Stress",
  p1pmsx = "Monitoring Social Experience"
)

# Loop through and generate pastel violin plots without legend
for (var in names(plot_vars)) {
  title <- plot_vars[[var]]
  p <- ggplot(df, aes(x = as.factor(p1emp2), y = .data[[var]], fill = as.factor(p1emp2))) +
    geom_violin(trim = FALSE) +
    geom_jitter(width = 0.2, alpha = 0.3, color = "black") +
    scale_fill_brewer(palette = "Pastel1") +
    labs(
      title = paste(title, "by Parent 2 Emp Status (", var, ")"),
      x = "Emp Status",
      y = title
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")
  print(p)
}
```

## ANOVA and Effect Size

### ANOVA vs Education

```{r}

df <- read.csv("w1-parent-scored.csv")
psychometrics <- list(
  p1sp   = "Parental Support",
  p1anxdp = "Anxiety/Depression",
  p1wtd  = "Weighted Stress",
  p1rbb  = "Risky Behavior Beliefs",
  p1agg  = "Aggression",
  p1smc  = "Social Media Concerns",
  p1thpr = "Thought Problems",
  p1atnpr = "Attention Problems",
  p1pix  = "Parental Involvement Index",
  p1ppx  = "Parental External Stress",
  p1pmsx = "Monitoring Social Experience"
)

# Loop through and print ANOVA summary for each
for (var in names(psychometrics)) {
  cat("\n--------------------------------------------\n")
  cat("ANOVA for", psychometrics[[var]], "(", var, ")\n")
  cat("--------------------------------------------\n")
  
  result <- summary(aov(as.formula(paste(var, "~ as.factor(p1edu1)")), data = df))
  print(result)
}

```

```{r}

df <- read.csv("w1-parent-scored.csv")
psychometrics <- list(
  p1sp   = "Parental Support",
  p1anxdp = "Anxiety/Depression",
  p1wtd  = "Weighted Stress",
  p1rbb  = "Risky Behavior Beliefs",
  p1agg  = "Aggression",
  p1smc  = "Social Media Concerns",
  p1thpr = "Thought Problems",
  p1atnpr = "Attention Problems",
  p1pix  = "Parental Involvement Index",
  p1ppx  = "Parental External Stress",
  p1pmsx = "Monitoring Social Experience"
)

# Loop through and print ANOVA summary for each
for (var in names(psychometrics)) {
  cat("\n--------------------------------------------\n")
  cat("ANOVA for", psychometrics[[var]], "(", var, ")\n")
  cat("--------------------------------------------\n")
  
  result <- summary(aov(as.formula(paste(var, "~ as.factor(p1edu2)")), data = df))
  print(result)
}
```

### ANOVA vs Marital Status

```{r}

df <- read.csv("w1-parent-scored.csv")
psychometrics <- list(
  p1sp   = "Parental Support",
  p1anxdp = "Anxiety/Depression",
  p1wtd  = "Weighted Stress",
  p1rbb  = "Risky Behavior Beliefs",
  p1agg  = "Aggression",
  p1smc  = "Social Media Concerns",
  p1thpr = "Thought Problems",
  p1atnpr = "Attention Problems",
  p1pix  = "Parental Involvement Index",
  p1ppx  = "Parental External Stress",
  p1pmsx = "Monitoring Social Experience"
)

# Loop through and print ANOVA summary for each
for (var in names(psychometrics)) {
  cat("\n--------------------------------------------\n")
  cat("ANOVA for", psychometrics[[var]], "(", var, ")\n")
  cat("--------------------------------------------\n")
  
  result <- summary(aov(as.formula(paste(var, "~ as.factor(p1mar1)")), data = df))
  print(result)
}
```

### ANOVA vs Employment

```{r}

df <- read.csv("w1-parent-scored.csv")
psychometrics <- list(
  p1sp   = "Parental Support",
  p1anxdp = "Anxiety/Depression",
  p1wtd  = "Weighted Stress",
  p1rbb  = "Risky Behavior Beliefs",
  p1agg  = "Aggression",
  p1smc  = "Social Media Concerns",
  p1thpr = "Thought Problems",
  p1atnpr = "Attention Problems",
  p1pix  = "Parental Involvement Index",
  p1ppx  = "Parental External Stress",
  p1pmsx = "Monitoring Social Experience"
)

# Loop through and print ANOVA summary for each
for (var in names(psychometrics)) {
  cat("\n--------------------------------------------\n")
  cat("ANOVA for", psychometrics[[var]], "(", var, ")\n")
  cat("--------------------------------------------\n")
  
  result <- summary(aov(as.formula(paste(var, "~ as.factor(p1emp1)")), data = df))
  print(result)
}
```

```{r}

df <- read.csv("w1-parent-scored.csv")
psychometrics <- list(
  p1sp   = "Parental Support",
  p1anxdp = "Anxiety/Depression",
  p1wtd  = "Weighted Stress",
  p1rbb  = "Risky Behavior Beliefs",
  p1agg  = "Aggression",
  p1smc  = "Social Media Concerns",
  p1thpr = "Thought Problems",
  p1atnpr = "Attention Problems",
  p1pix  = "Parental Involvement Index",
  p1ppx  = "Parental External Stress",
  p1pmsx = "Monitoring Social Experience"
)

# Loop through and print ANOVA summary for each
for (var in names(psychometrics)) {
  cat("\n--------------------------------------------\n")
  cat("ANOVA for", psychometrics[[var]], "(", var, ")\n")
  cat("--------------------------------------------\n")
  
  result <- summary(aov(as.formula(paste(var, "~ as.factor(p1edu2)")), data = df))
  print(result)
}
```

### Effect Size and Kruskal-Wallis vs Education

```{r}
library(effectsize)

# Load data
data <- read.csv("w1-parent-scored.csv")

# Psychometric variables
psychometrics <- list(
  p1sp   = "Parental Support",
  p1anxdp = "Anxiety/Depression",
  p1wtd  = "Weighted Stress",
  p1rbb  = "Risky Behavior Beliefs",
  p1agg  = "Aggression",
  p1smc  = "Social Media Concerns",
  p1thpr = "Thought Problems",
  p1atnpr = "Attention Problems",
  p1pix  = "Parental Involvement Index",
  p1ppx  = "Parental External Stress",
  p1pmsx = "Monitoring Social Experience"
)

# Loop: run ANOVA and compute eta squared
for (var in names(psychometrics)) {
  formula <- as.formula(paste(var, "~ as.factor(p1edu1)"))
  
  # Run ANOVA
  model <- aov(formula, data = data)
  
  # Eta Squared
  eta <- eta_squared(model, partial = FALSE)
  
  # Kruskal-Wallis
  kw_p <- kruskal.test(formula, data = data)$p.value
  
  cat("\n----------------------------------------\n")
  cat("Results for", psychometrics[[var]], "(", var, ")\n")
  cat("----------------------------------------\n")
  print(eta)
  cat("Kruskal-Wallis p-value:", round(kw_p, 4), "\n")
}

```

```{r}
library(effectsize)

# Load data
data <- read.csv("w1-parent-scored.csv")

# Psychometric variables
psychometrics <- list(
  p1sp   = "Parental Support",
  p1anxdp = "Anxiety/Depression",
  p1wtd  = "Weighted Stress",
  p1rbb  = "Risky Behavior Beliefs",
  p1agg  = "Aggression",
  p1smc  = "Social Media Concerns",
  p1thpr = "Thought Problems",
  p1atnpr = "Attention Problems",
  p1pix  = "Parental Involvement Index",
  p1ppx  = "Parental External Stress",
  p1pmsx = "Monitoring Social Experience"
)

# Loop: run ANOVA and compute eta squared
for (var in names(psychometrics)) {
  formula <- as.formula(paste(var, "~ as.factor(p1edu2)"))
  
  # Run ANOVA
  model <- aov(formula, data = data)
  
  # Eta Squared
  eta <- eta_squared(model, partial = FALSE)
  
  # Kruskal-Wallis
  kw_p <- kruskal.test(formula, data = data)$p.value
  
  cat("\n----------------------------------------\n")
  cat("Results for", psychometrics[[var]], "(", var, ")\n")
  cat("----------------------------------------\n")
  print(eta)
  cat("Kruskal-Wallis p-value:", round(kw_p, 4), "\n")
}

```

### Effect Size and Kruskal-Wallis vs Marital Status

```{r}
library(effectsize)

# Load data
data <- read.csv("w1-parent-scored.csv")

# Psychometric variables
psychometrics <- list(
  p1sp   = "Parental Support",
  p1anxdp = "Anxiety/Depression",
  p1wtd  = "Weighted Stress",
  p1rbb  = "Risky Behavior Beliefs",
  p1agg  = "Aggression",
  p1smc  = "Social Media Concerns",
  p1thpr = "Thought Problems",
  p1atnpr = "Attention Problems",
  p1pix  = "Parental Involvement Index",
  p1ppx  = "Parental External Stress",
  p1pmsx = "Monitoring Social Experience"
)

# Loop: run ANOVA and compute eta squared
for (var in names(psychometrics)) {
  formula <- as.formula(paste(var, "~ as.factor(p1mar1)"))
  
  # Run ANOVA
  model <- aov(formula, data = data)
  
  # Eta Squared
  eta <- eta_squared(model, partial = FALSE)
  
  # Kruskal-Wallis
  kw_p <- kruskal.test(formula, data = data)$p.value
  
  cat("\n----------------------------------------\n")
  cat("Results for", psychometrics[[var]], "(", var, ")\n")
  cat("----------------------------------------\n")
  print(eta)
  cat("Kruskal-Wallis p-value:", round(kw_p, 4), "\n")
}

```

### Effect Size and Kruskal-Wallis vs Employment

```{r}
library(effectsize)

# Load data
data <- read.csv("w1-parent-scored.csv")

# Psychometric variables
psychometrics <- list(
  p1sp   = "Parental Support",
  p1anxdp = "Anxiety/Depression",
  p1wtd  = "Weighted Stress",
  p1rbb  = "Risky Behavior Beliefs",
  p1agg  = "Aggression",
  p1smc  = "Social Media Concerns",
  p1thpr = "Thought Problems",
  p1atnpr = "Attention Problems",
  p1pix  = "Parental Involvement Index",
  p1ppx  = "Parental External Stress",
  p1pmsx = "Monitoring Social Experience"
)

# Loop: run ANOVA and compute eta squared
for (var in names(psychometrics)) {
  formula <- as.formula(paste(var, "~ as.factor(p1emp1)"))
  
  # Run ANOVA
  model <- aov(formula, data = data)
  
  # Eta Squared
  eta <- eta_squared(model, partial = FALSE)
  
  # Kruskal-Wallis
  kw_p <- kruskal.test(formula, data = data)$p.value
  
  cat("\n----------------------------------------\n")
  cat("Results for", psychometrics[[var]], "(", var, ")\n")
  cat("----------------------------------------\n")
  print(eta)
  cat("Kruskal-Wallis p-value:", round(kw_p, 4), "\n")
}

```

```{r}
library(effectsize)

# Load data
data <- read.csv("w1-parent-scored.csv")

# Psychometric variables
psychometrics <- list(
  p1sp   = "Parental Support",
  p1anxdp = "Anxiety/Depression",
  p1wtd  = "Weighted Stress",
  p1rbb  = "Risky Behavior Beliefs",
  p1agg  = "Aggression",
  p1smc  = "Social Media Concerns",
  p1thpr = "Thought Problems",
  p1atnpr = "Attention Problems",
  p1pix  = "Parental Involvement Index",
  p1ppx  = "Parental External Stress",
  p1pmsx = "Monitoring Social Experience"
)

# Loop: run ANOVA and compute eta squared
for (var in names(psychometrics)) {
  formula <- as.formula(paste(var, "~ as.factor(p1emp2)"))
  
  # Run ANOVA
  model <- aov(formula, data = data)
  
  # Eta Squared
  eta <- eta_squared(model, partial = FALSE)
  
  # Kruskal-Wallis
  kw_p <- kruskal.test(formula, data = data)$p.value
  
  cat("\n----------------------------------------\n")
  cat("Results for", psychometrics[[var]], "(", var, ")\n")
  cat("----------------------------------------\n")
  print(eta)
  cat("Kruskal-Wallis p-value:", round(kw_p, 4), "\n")
}

```

## Unsupervised Learning
